\input{picture}

\section{Introduction}
\begin{frame}
  \frametitle{Mutalyzer: a curational tool for \emph{Locus Specific Mutation
    Databases (LsdBs)}}
  \medskip

  Variant nomenclature checker applying \emph{Human Genome Variation Society}
  (HGVS) guidelines.
  \medskip
  \pause
  \begin{itemize}
    \item Is the syntax of the variant description valid?
    \item Does the reference sequence exist?
    \item Is the variant possible on this reference sequence?
    \item Is this variant description the recommended one?
  \end{itemize}
  \bigskip
  \medskip
  \pause

  Basic effect prediction.
  \medskip
  \begin{itemize}
    \item Is the description of the transcript product as expected?
    \item Is the predicted protein as expected?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{HGVS nomenclature}

  \emph{Genomic} orientated positions:
  \begin{center}
    \bt{AL449423.14:g.[65449\_65463del;65564\color{yellow}T\color{white}>\color{yellow}C\color{white}]}
  \end{center}
  \pause
  \bigskip
  \emph{Coding sequence} orientated positions:
  \begin{center}
    \bt{AL449423.14(CDKN2A\_v001):c.[5\color{yellow}A\color{white}>\color{yellow}G\color{white}
      ;106\_120del]}
  \end{center}
  \bigskip
  \pause
  \begin{itemize}
    \item \bt{AL449423.14} -- reference sequence.
    \item \bt{CDKN2A\_v001}$\;$ -- transcript variant \bt{1} of gene CDKN2A.
    \item \bt{c.[5\color{yellow}A\color{white}>\color{yellow}G\color{white};106\_120del]}
    \begin{itemize}
      \item A \emph{substitution} at position \bt{5} counting from the start
        codon.
      \item A \emph{deletion} from position \bt{106} to position \bt{120}.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Coordinate systems}

  \setlength{\unitlength}{1pt}
  \positionpicture

  \renewcommand{\arraystretch}{1}
  \begin{center}
    \begin{tabular}{l|r|r|r}
      Name                              & \bt{g.}  & \bt{n.}      & \bt{c.} \\
      \hline
      {\scriptsize Genomic start}       & \bt{1}   & \bt{100+d70} &
        \bt{*10+d70} \\
      {\scriptsize Genomic end}         & \bt{300} & \bt{1-u50}   &
        \bt{-30-u50} \\
      {\scriptsize Transcription start} & \bt{250} & \bt{1}       & \bt{-30} \\
      {\scriptsize Transcription end}   & \bt{70}  & \bt{100}     & \bt{*10} \\
      {\scriptsize CDS start}           & \bt{220} & \bt{30}      & \bt{1} \\
      {\scriptsize CDS stop}            & \bt{80}  & \bt{90}      & \bt{60} \\
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}
  \frametitle{Coordinate systems}

  \setlength{\unitlength}{1pt}
  \positionpicture

  \bt{c.} positions:
  \begin{itemize}
    \item Positions in introns are relative to the nearest exonic position.
    \item Positions before the CDS are indicated with a \bt{-} sign.
    \item Positions after the CDS are indicated with a \bt{*} sign.
    \medskip
    \pause
    \item Position \bt{-1} and \bt{1} are adjacent.
    \item If \bt{60} is the last position of the CDS, then \bt{60} and \bt{*1}
      are adjacent.
  \end{itemize}
\end{frame}
