\begin{longtable}{llp{7cm}}
name     & $\rightarrow$ & ([a--z][A--Z][0--9])$^+$\\
Nt       & $\rightarrow$ & `a' $|$ `c' $|$ `g' $|$ `t' $|$ `u' $|$ `A' $|$ 
                           `C' $|$ `G' $|$ `T' $|$ `U'\\
NtString & $\rightarrow$ & (Nt)$^+$\\
Number & $\rightarrow$ & ([0--9])$^+$\\
TransVar   & $\rightarrow$ & `\_v' Number\\
ProtIso    & $\rightarrow$ & `\_i' Number\\
GeneSymbol & $\rightarrow$ & `(' name (TransVar $|$ ProtIso)? `)'\\
GI         & $\rightarrow$ & (`GI' $|$ `GI:')? Number\\
Version    & $\rightarrow$ & `.' Number\\
AccNo      & $\rightarrow$ & `ish.'? ([a--Z] Number `\_')$^+$ Version? \\
UD         & $\rightarrow$ & `UD\_' name (`\_' Number)$^+$\\
LRGTranscriptID & $\rightarrow$ & `t' ([0-9])$^+$\\
LRGProteinID    & $\rightarrow$ & `p' ([0-9])$^+$\\
LRG             & $\rightarrow$ & `LRG' ([0-9])$^+$ (`\_' (LRGTranscriptID $|$ 
                  LRGProteinID))?\\
GenBankRef  & $\rightarrow$ & (GI $|$ AccNo $|$ UD) (`(' GeneSymbol `)')?\\
RefSeqAcc & $\rightarrow$ & GenBankRef $|$ LRG\\
Chrom & $\rightarrow$ & name\\
Offset & $\rightarrow$ & (`+' $|$ `-') (`u' $|$ `d')? (Number $|$ `?')\\
RealPtLoc & $\rightarrow$ & ((`-' $|$ `*')? Number Offset?) $|$ `?'\\
IVSLoc & $\rightarrow$ & `IVS' Number (`+' $|$ `-') Number\\
PtLoc  & $\rightarrow$ & IVSLoc $|$ RealPtLoc\\
RefType & $\rightarrow$ & (`c' $|$ `g' $|$ `m' $|$ `n' $|$ `r') `.'\\
Ref & $\rightarrow$ & ((RefSeqAcc $|$ GeneSymbol) `:')? RefType?\\
RealExtent & $\rightarrow$ & PtLoc `\_' (`o'? (RefSeqAcc $|$ GeneSymbol) `:')? PtLoc\\
EXLoc & $\rightarrow$ & `EX' Number (`-' Number)?\\
Extent & $\rightarrow$ & RealExtent $|$ EXLoc\\
RangeLoc & $\rightarrow$ & Extent $|$ `(` Extent `)'\\
Loc & $\rightarrow$ & PtLoc $|$ RangeLoc\\
FarLoc & $\rightarrow$ & (RefSeqAcc $|$ GeneSymbol) (`:' RefType? Extent)? \\
Subst & $\rightarrow$ & PtLoc Nt `$>$' Nt\\
Del & $\rightarrow$ & Loc `del' (Nt$^+$ $|$ Number)?\\
Dup & $\rightarrow$ & Loc `dup' (Nt$^+$ $|$ Number)? Nest?\\
AbrSSR & $\rightarrow$ & PtLoc  Nt$^+$ `(' Number `\_' Number `)'\\
VarSSR & $\rightarrow$ & (PtLoc  Nt$^+$ `[' Number `]') $|$ (RangeLoc `[' Number `]') \\
Ins & $\rightarrow$ & RangeLoc `ins' (Nt$^+$ $|$ Number $|$ RangeLoc $|$ FarLoc) Nest?\\
Indel & $\rightarrow$ & RangeLoc `del' (Nt$^+$ $|$ Number)? `ins' (Nt$^+$ $|$ Number $|$ RangeLoc $|$ FarLoc) Nest?\\
Inv & $\rightarrow$ & RangeLoc `inv' (Nt$^+$ $|$ Number)? Nest?\\
Conv & $\rightarrow$ & RangeLoc `con' FarLoc Nest?\\
ChromBand & $\rightarrow$ & (`p' $|$ `q') Number `.' Number\\
ChromCoords & $\rightarrow$ & `(' Chrom `;' Chrom `)' `(' ChromBand `;' ChromBand `)'\\
TransLoc & $\rightarrow$ & `t' ChromCoords `(' FarLoc `)'\\
RawVar & $\rightarrow$ & Subst $|$ Del $|$ Dup $|$ VarSSR $|$ Ins $|$ Indel $|$ Inv $|$ Conv\\
SingleVar & $\rightarrow$ & Ref RawVar $|$ TransLoc\\
ExtendedRawVar & $\rightarrow$ & RawVar $|$ `=' $|$ `?'\\
CAlleleVarSet & $\rightarrow$ & ExtendedRawVar (`;' ExtendedRawVar)$^*$\\
UAlleleVarSet & $\rightarrow$ & (CAlleleVarSet $|$ (`(' CAlleleVarSet `)')) `?'?\\
SimpleAlleleVarSet & $\rightarrow$ & (`[' UAlleleVarSet `]') $|$ ExtendedRawVar\\
MosaicSet & $\rightarrow$ & (`[' SimpleAlleleVarSet (`/' SimpleAlleleVarSet)$^*$ `]') $|$ SimpleAlleleVarSet\\
ChimeronSet & $\rightarrow$ & (`[' MosaicSet (`//' MosaicSet)$^*$ `]') $|$ MosaicSet\\
SingleAlleleVarSet & $\rightarrow$ & (`[' ChimeronSet ((`;' $|$ `\verb#^#') ChimeronSet)$^*$ (`(;)' ChimeronSet)$^*$ `]') $|$ ChimeronSet\\
SingleAlleleVars & $\rightarrow$ & Ref SingleAlleleVarSet\\
MultiAlleleVars & $\rightarrow$ & Ref SingleAlleleVarSet (`;' Ref? SingleAlleleVarSet)$^+$\\
MultiVar & $\rightarrow$ & SingleAlleleVars $|$ MultiAlleleVars\\
MultiTranscriptVar & $\rightarrow$ & Ref `[` ExtendedRawVar (`;' ExtendedRawVar)$^*$ (`,' ExtendedRawVar (`;' ExtendedRawVar)$^*$)$^+$ `]'\\
UnkEffectVar & $\rightarrow$ & Ref (`(=)' $|$ `?')\\
SplicingVar & $\rightarrow$ & Ref (`spl?' $|$ `(spl?)')\\
NoRNAVar & $\rightarrow$ & Ref `0' `?'?\\
Var & $\rightarrow$ & SingleVar $|$ MultiVar $|$ MultiTranscriptVar $|$ UnkEffectVar $|$ NoRNAVar $|$ SplicingVar\\
Nest & $\rightarrow$ & `\{' SimpleAlleleVarSet `\}'
\end{longtable}
